---
title: "Hayman Fire Recovery"
author: "Derek Fedak"
date: "`r format(Sys.time(), '%A %d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, warning=F,message=F}
library(tidyverse)
library(tidyr)
library(ggthemes)
library(gridExtra)
library(lubridate)

# Now that we have learned how to munge (manipulate) data
# and plot it, we will work on using these skills in new ways

knitr::opts_knit$set(root.dir='..')
```


```{r dataread, warning=F,message=F}
####-----Reading in Data and Stacking it ----- ####
#Reading in files
files <- list.files('data',full.names=T)


#Read in individual data files
ndmi <- read_csv(files[1]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')


ndsi <- read_csv(files[2]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')

ndvi <- read_csv(files[3])%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')

# Stack as a tidy dataset
full_long <- rbind(ndvi,ndmi,ndsi) %>%
  gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value))

# Example stack as a tidy dataset
full_long2 <- rbind(ndvi, ndmi, ndsi) %>%
  gather(key='treatment', value='veg_index', -DateTime, -data) %>%
  filter(!is.na('veg_index'))

```


## Question 1) 

What is the correlation between NDVI and NDMI? - here I want you to
convert the full_long dataset in to a wide dataset using the 
function "spread" and then make a plot that shows the correlation s a
function of if the site was burned or not (x axis should be ndmi)
You should exclude winter months and focus on summer months

### Answer 1
There is a positive correlation between NDVI and NDMI, which makes intuitive
sense that higher moisture may correlate to increased vegetation 'greenness'.
Isolating for summer months while separating into burn classes also highlights
that both NDVI and NDMI are greater in the unburned areas when compared to the
burned areas.

```{r}
q <- pivot_wider(full_long, names_from = "data", values_from = "value") %>%
  filter_if(is.numeric, all_vars(!is.na(.))) %>%
  mutate(month=month(DateTime),
       year=year(DateTime))
q_summer <- filter(q, month %in% c(6,7,8,9))

ggplot(q_summer,aes(x=ndmi,y=ndvi,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.8))

```


## Question 2 

2) What is the correlation between average NDSI (normalized
 snow index) for January - April and average NDVI for June-August?
In other words, does the previous year's snow cover influence vegetation
 growth for the following summer?
 
### Answer 2
There may be a very slight but weak positive correlation between previous
winter's NDSI and current summer's NDVI. NDVI typically ranges >0.38 for both
unburned sites and burned sites pre-fire. The clustering of NDVI values <0.3
represents post-fire burned sites.

```{r}
q_annual_ndsi <- q %>%
  mutate(year=year(DateTime)) %>%
  mutate(month=month(DateTime)) %>%
  filter(month %in% c(1,2,3,4)) %>%
  group_by(site,year) %>%
  summarize(mean_NDSI=mean(ndsi))

q_annual_ndvi <- q %>%
  mutate(year=year(DateTime)) %>%
  mutate(month=month(DateTime)) %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(site,year) %>%
  summarize(mean_NDVI=mean(ndvi))

q_snow2veg <- left_join(q_annual_ndsi, q_annual_ndvi)

ggplot(q_snow2veg,aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.2))

```

## Q3

How is the snow effect from question 2 different between pre- and post-burn
and burned and unburned?

### Answer 3
We can see a pretty drastic shift in post-fire NDVI between burned and unburned
areas. We see similar mean NDSI values, but unburned NDVI is greater in all of
these cases when compared to the NDVI in burned locations. This trend is not
noticeable in the pre-fire values where the yet-to-be-burned locations have
greater NDVI values consistent with the unburned locations.

```{r}
q_snow2veg_burn <- q_snow2veg %>% filter(site=="burned")
q_snow2veg_unburn <- q_snow2veg %>% filter(site=="unburned")

#Breaking out burned/unburned entirely into different charts of pre- and post-fire
p1 <- ggplot(q_snow2veg_unburn %>% filter(year<2002),aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  ggtitle("Pre-fire, Unburned")

p2 <- ggplot(q_snow2veg_unburn %>% filter(year>2002),aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  ggtitle("Post-fire, Unburned")

p3 <- ggplot(q_snow2veg_burn %>% filter(year<2002),aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  ggtitle("Pre-fire, Burned")

p4 <- ggplot(q_snow2veg_burn %>% filter(year>2002),aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  ggtitle("Post-fire, Burned")
grid.arrange(p1, p2, p3, p4)

#Combining burned/unburned but still separated by pre- and post-fire. More clean, better comparison
p1 <- ggplot(q_snow2veg %>% filter(year<2002),aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  ggtitle("Pre-fire (1984-2001)")

p2 <- ggplot(q_snow2veg %>% filter(year>2002),aes(x=mean_NDSI,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  ggtitle("Post-fire (2003-2019)")
grid.arrange(p1,p2)
```


## Question 4

What month is the greenest month on average? 

### Answer 4

The greenest month on average for all sites and years is August with an average
NDVI of 0.387, while September has an average NDVI of 0.383. When we analyze
just unburned locations across the time period, September is the greenest month
with an average NDVI of 0.459, though June-October have average NDVI > 0.4. 

```{r}
#All sites across time
q_monthly_ndvi <- q %>%
  mutate(month=month(DateTime)) %>%
  group_by(month) %>%
  summarize(mean_NDVI=mean(ndvi))

#Broken out by site
q_monthly_ndvi2 <- q %>%
  mutate(month=month(DateTime)) %>%
  group_by(month, site) %>%
  summarize(mean_NDVI=mean(ndvi))

#Combining burned/unburned but still separated by pre- and post-fire. More clean, better comparison
ggplot(q_monthly_ndvi2,aes(x=month,y=mean_NDVI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  scale_x_continuous(breaks=seq(0,12,by=1)) +
  ggtitle("NDVI by Month")

```



## Question 5

What month is the snowiest on average?

### Answer 5

The snowiest month on average for all sites and years is January with an average
NDSI of 0.210, while February has an average NDSI of 0.198. When we analyze
just unburned locations across the time period, February is the snowiest month
with an average NDSI of 0.273, and December-February is consistently snowier in
unburned sites than burned sites.

```{r}
#All sites across time
q_monthly_ndsi <- q %>%
  mutate(month=month(DateTime)) %>%
  group_by(month) %>%
  summarize(mean_NDSI=mean(ndsi))

#Broken out by site
q_monthly_ndsi2 <- q %>%
  mutate(month=month(DateTime)) %>%
  group_by(month, site) %>%
  summarize(mean_NDSI=mean(ndsi))

#Combining burned/unburned but still separated by pre- and post-fire. More clean, better comparison
ggplot(q_monthly_ndsi2,aes(x=month,y=mean_NDSI,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  scale_x_continuous(breaks=seq(0,12,by=1)) +
  ggtitle("NDSI by Month")
```

## Bonus Question

Redo all problems with `spread` and `gather` using modern tidyverse syntax. 


## Bonus Question

Use Climate Engine to pull the same data for the assignment, but updated with 2020/2021 data.




